<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bettingtools

Author: Peter Wu

<!-- badges: start -->
<!-- badges: end -->

The bettingtools package has functions related to working with sports betting lines. First, we provide functions to work with American, Decimal, and Implied odds in the tidy format. Next, we provide functions to work with sports betting lines such as calculating zero-vig probabilities, theoretical hold, and the optimal single Kelly stake bet. Furthermore, the package comes with tidy datasets for historical NBA (work in progress), MLB, NFL (work in progress), and NHL (work in progress) Vegas lines for moneylines, point totals, and point spreads.

These functions take inspiration from posts by user "Ganchrow" on the Sportsbook Review forum.

Link to [original GitHub respiratory](https://github.com/pwu97/bettingtools)

Link to deployed website

Customizations:

1.  Added template
2.  Added links to sidebar
3.  Modified footer
4.  Modified navigation bar height
5.  Added github link to navbar

## Installation

``` r
remotes::install_github("pwu97/bettingtools")
```

## Exported Functions and Examples 

We calculate the average over/under movement, the average away/home closing moneyline, and whether the closing moneylines are sharper than the opening moneylines given by Vegas, on average. We can see that home teams tended to be favored more often, on average, than away teams. Furthermore, there is also evidence that closing moneylines were indeed sharper than opening moneylines by about 2%.

```{r}
library(bettingtools)
library(teamcolors)
library(here)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(forcats)
library(usethis)
library(readr)
library(purrr)

# What is the average O/U movement for games in which there was line movement?
mlb_odds_2019 %>%
  mutate(ou_movement = close_ou_line - open_ou_line) %>%
  filter(ou_movement != 0) %>%
  summarize(mean_ou_movement = mean(ou_movement)) %>%
  pull(mean_ou_movement)

# What is the average away closing moneyline?
mlb_odds_2019 %>%
  mutate(away_implied = US2Implied(away_close_ml)) %>%
  summarize(avg_away_closing_ml = Implied2US(mean(away_implied))) %>%
  pull(avg_away_closing_ml)

# What is the average home closing moneyline?
mlb_odds_2019 %>%
  mutate(home_implied = US2Implied(home_close_ml)) %>%
  summarize(avg_home_closing_ml = Implied2US(mean(home_implied))) %>%
  pull(avg_home_closing_ml)

# Are closing moneylines sharper than opening moneylines?
mlb_odds_2019 %>%
  mutate(fav_team_won_open = ifelse((((away_score > home_score) & 
                                      (away_open_ml < home_open_ml)) |
                                     ((away_score < home_score) & 
                                      (away_open_ml > home_open_ml))), 1, 0),
         fav_team_won_close = ifelse((((away_score > home_score) & 
                                      (away_close_ml < home_close_ml)) |
                                     ((away_score < home_score) & 
                                      (away_close_ml > home_close_ml))), 1, 0)) %>%
  summarize(pct_fav_won_open = mean(fav_team_won_open, na.rm = TRUE),
            pct_fav_won_close = mean(fav_team_won_close, na.rm = TRUE),
            pct_diff = pct_fav_won_close - pct_fav_won_open)
```

### Calculate single Kelly stake

We can calculate the percentage of one's bankroll one should bet to maximize the expected growth of one's bankroll on a single bet. Given an expected win probability, payout odds, and an optional Kelly multiplier factor, we can calculate one's optimal single Kelly stake. Note that default odds for the expected win probability is implied probability ("prob") and the default odds for the payout is in decimal ("dec"). We can change them accordingly to how we see fit by specifying additional parameters in our function ("prob", "dec", "us").

```{r example}
calculateKellyStake(0.53, 1.92)

calculateKellyStake(0.41, 2.56)

# Getting a bet at +150 when it is expected to hit at -120
calculateKellyStake(-120, 150, expected_odds = "us", payout_odds = "us")

# Getting a bet at -150 when expected probability is 70% at a 10% kelly multiplier.
calculateKellyStake(0.70, -150, kelly_multiplier = 0.1, payout_odds = "us")

# Optimal move is to not place a bet
calculateKellyStake(0.26, -110, payout_odds = "us")
```

### Calculate all possible win-loss outcomes given probabilities

We return a tibble where each row is one possible outcome we can observe from a list of given probabilities.

```{r}
calculateWinRanges(c(.1, .4, .88, .47))

calculateWinRanges(c(.12, .462, .29))

calculateWinRanges(c(.6, .6, .6, .6))
```


### Calculate zero-vig implied probabilities

We can calculate the zero-vig implied probabilities of a vector of lines. The default precision is set to 4 digits. Note that we can set the precision.

```{r}
calculateZeroVigProb(c(200, -180, -450, 800))

calculateZeroVigProb(-237)

calculateZeroVigProb(-237, precision = 7)
```

###  Calculate implied probabilities for two-outcome line set

We can calculate the implied probabilities for two or more lines by first calculating the zero-vig implied probabilities for both of them, and then normalizing them. Again, we can set the precision. 

```{r}
calculateNormalizedImplied(c(1000, -800), precision = 7)

calculateNormalizedImplied(c(200, -220))

calculateNormalizedImplied(c(427, -213, 336))
```

### Calculate theoretical hold

We can calculate the theoretical hold for a two-outcome line set. This corresponds to the profit a sportsbook would expect to make were a player to bet on either side of an event with all else being equal. Contrary to popular belief, larger nominal spreads doesn't necessarily mean more profit for bookies.

```{r}
calculateTheoreticalHold(c(-110, -110))

calculateTheoreticalHold(c(-1500, 875))

calculateTheoreticalHold(c(-1500, 875), precision = 7)
```

### Converting between American, Decimal, and Implied Odds

We can convert between American, Decimal, and Implied Odds.

```{r}
US2Implied(c(-250, 600, 137, -110))

US2Dec(c(-250, 600, 137, -110))

US2All(c(-250, 600, 137, -110))

Dec2Implied(c(3.17, 2.14, 2.01, 1.67))

Dec2US(c(3.17, 2.14, 2.01, 1.67))

Dec2All(c(3.17, 2.14, 2.01, 1.67))

Implied2Dec(c(.34, .54, .88, .12))

Implied2US(c(.34, .54, .88, .12))

Implied2All(c(.34, .54, .88, .12))
```

