---
title: "vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Here, we'll show how to use the functions in the package for an exemplary data analysis. Our goal is to see the relationship between pre-game favorites and end-game results.

The data we will analyze is about MLB and NBA betting odds.

Link to [data](https://sports-statistics.com/sports-data/)

## Data Dictionary:

### MLB Data

| variable           | class     | description                                          |
|:----------------------|:------------------|:-----------------------------|
| date               | character | date of the game                                     |
| game               | integer   | game number                                          |
| away_abbrev        | character | abbreviation of away team name                       |
| home_abbrev        | character | abbreviation of home team name                       |
| away_name          | character | away team name                                       |
| home_name          | character | home team name                                       |
| away_score         | integer   | final away team score                                |
| home_score         | integer   | final home team score                                |
| away_open_ml       | integer   | first away team moneyline                            |
| home_open_ml       | integer   | first home team moneyline                            |
| away_run_line      | double    | away team run line                                   |
| home_run_line      | double    | home team run line                                   |
| away_run_line_odds | integer   | odds for away team run line                          |
| home_run_line_odds | integer   | odds for home team run line                          |
| open_ou_line       | double    | first Over/Under line for total runs scored          |
| open_ou_odds       | integer   | odds for first Over/Under line for total runs scored |
| close_ou_line      | double    | last Over/Under line for total runs scored           |
| close_ou_odds      | integer   | odds for last Over/Under line for total runs scored  |

### NBA Data

| variable   | class     | description           |
|:-----------|:----------|:----------------------|
| date       | character | date of the game      |
| game       | integer   | game number           |
| away_name  | character | away team name        |
| home_name  | character | home team name        |
| away_score | integer   | final away team score |
| home_score | integer   | final home team score |
| away_ml    | integer   | away team moneyline   |
| home_ml    | integer   | home team moneyline   |

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(bettingtools)
library(teamcolors)
library(here)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(forcats)
library(usethis)
library(readr)
library(purrr)
```

```{r, warning=FALSE, message=FALSE}
if (!dir.exists(here("data"))) {
    dir.create(here("data"))
}
```

```{r}
mlb_colors <- teamcolors::teamcolors %>%
  filter(league == "mlb") %>%
  rename(team = name) %>%
  select(team, primary)

away_team_lines <- mlb_odds_2019 %>%
  select(away_name, away_close_ml) %>%
  mutate(implied_ml = US2Implied(away_close_ml)) %>%
  rename(team = away_name) %>%
  select(team, implied_ml)

home_team_lines <- mlb_odds_2019 %>%
  select(home_name, home_close_ml) %>%
  mutate(implied_ml = US2Implied(home_close_ml)) %>%
  rename(team = home_name) %>%
  select(team, implied_ml)

avg_2019_mlb_ml <- rbind(away_team_lines, home_team_lines) %>%
  group_by(team) %>%
  summarize(avg_implied_ml = mean(implied_ml),
            avg_ml_line = Implied2US(mean(implied_ml))) %>%
  left_join(mlb_colors, by = "team") %>%
  arrange(desc(avg_ml_line)) %>%
  slice(-n()) %>%
  mutate(team = fct_reorder(team, avg_ml_line, .desc = TRUE),
         primary = fct_reorder(primary, avg_ml_line, .desc = TRUE)) %>%
  ungroup()

avg_2019_mlb_ml %>%
  ggplot(aes(x = team, y = avg_ml_line, label = avg_ml_line)) +
  geom_col(fill = avg_2019_mlb_ml$primary) +
  coord_flip() +
  labs(y = "Average Moneyline",
       x = "Team Name",
       title = "Average moneyline for each team during\nthe 2019 MLB Season",
       subtitle = "Houston Astros has the lowest ML on average",
       caption = "Data Source: Sports-Statistics.com ")

```

```{r}
nba_odds_2020 <- nba_odds_2020 %>%
  mutate(date = ymd(date))

nba_odds_2020 <- nba_odds_2020 %>%
  mutate(
    game_result = purrr::map2(home_score, away_score, ~ case_when(
      .x > .y ~ "Home Win",
      .x < .y ~ "Away Win",
      TRUE ~ "Tie"
    ))
  )

nba_odds_2020$game_result <- as.character(nba_odds_2020$game_result)

ggplot(nba_odds_2020, aes(x = home_ml, y = away_ml, color = game_result)) +
  geom_point() +
  labs(title = "Moneyline Distribution",
       subtitle = "NBA 2020 Season",
       x = "Home Moneyline",
       y = "Away Moneyline",
       caption = "Data Source: Sports-Statistics.com ")

```

```{r}
nba_odds_2020 <- nba_odds_2020 %>%
  mutate(higher_ml_winner = pmap_chr(list(home_ml, away_ml, game_result),
                                     function(home_ml, away_ml, game_result) {
                                       if ((game_result == "Home Win" && home_ml > away_ml) ||
                                           (game_result == "Away Win" && away_ml > home_ml)) {
                                         return("Yes")
                                       } else {
                                         return("No")
                                       }
                                     }))

ggplot(nba_odds_2020, aes(x = higher_ml_winner, fill = game_result)) +
  geom_bar() +
  labs(title = "Relationship Between Higher Moneyline and Game Result",
       subtitle = "NBA 2020 Season",
       x = "Higher Moneyline Winner",
       y = "Count",
       caption = "Data Source: Sports-Statistics.com ") +
  scale_fill_manual(values = c("Home Win" = "skyblue", "Away Win" = "coral"))


```

```{r}
nba_odds_2020$game_number <- seq_along(nba_odds_2020$away_name)
ggplot(nba_odds_2020, aes(x = game_number, y = away_ml, group = 1)) +
  geom_line(color = "blue", linetype = "dashed", aes(y = home_ml)) +
  facet_wrap(~game_result) +
  labs(title = "Moneyline Odds Across Games",
       subtitle = "NBA 2020 Season",
       x = "Game Number",
       y = "Moneyline Odds",
       caption = "Data Source: Sports-Statistics.com ")

```


## Summary

Moneyline for two teams are inversely correlated, with the lower odds moneyline having greater magnitude. "Underdog" away teams prevails more often than underdog home teams.

## Functions

dplyr: filter, mutate, group, select, left_join, arrange

lubridate: year

purrr: pmap, map2

ggplot2: geom_col, geom_point, geom_line, facet_wrap, geom_bar()
